---
title: "Disease"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{disease}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(dplyr)

options(scipen = 999)
```

## Introduction

Infectious disease models depend heavily on **accurate parameterization** to reflect real-world dynamics. Parameters such as the incubation period, infectious duration, case fatality rates, and levels of maternal protection all influence how a disease spreads and what burden it imposes. These values vary between diseases—and sometimes even between settings—so the ability to **customize disease-specific parameters** is critical for model realism.

The **PREVAIL** package allows users to flexibly define disease-specific characteristics using simple, human-readable data frames. This vignette shows you how to do that, whether you're:

- Replicating known case trends with historical WHO data  
- Adjusting natural history parameters (e.g. incubation or recovery rates)  
- Overriding default assumptions with new literature or expert input  
- Exploring hypothetical changes in severity or immunity for scenario analysis

By providing structured input objects (`custom_disease_parameters` and `custom_disease_data`), you can tailor the model to better match your disease of interest—while retaining the ability to fall back on PREVAIL’s validated defaults when needed.

You’ll learn how to:

- Define and structure disease-specific parameters  
- Supply custom case time series for dynamic or seeded simulations  
- Generate model-ready input using `custom_data_process_wrapper()`  
- Run and visualize simulations with your customized setup

This approach ensures transparency and reproducibility while giving you full control over how disease dynamics are represented in the model.

## 1. Prepare Your Custom Data

Before customizing the disease-specific behaviour of the transmission model, we need to load the required packages and understand the types of inputs we’ll be working with.

In this vignette, we focus on **disease parameters**—the core inputs that define how a pathogen behaves in the model. These include values like the incubation period, infectious duration, and case fatality rates, as well as assumptions about immunity and maternal protection. PREVAIL provides sensible defaults for several diseases, but users can override these with custom values as needed.

We also show how to provide a **custom case time series**, which can be used to seed the model based on known outbreaks or to constrain transmission to match historical data.

To get started, we load the required packages using the `pacman` package. This will install any missing packages and then load them into the session:

```{r, eval = FALSE}
# Install pacman if not already available
if (!require("pacman")) install.packages("pacman")

# Load required packages (installing if needed)
pacman::p_load(
  PREVAIL,     # The core transmission model and input processing framework
  tidyverse    # Tidy tools for data wrangling and plotting
)

```

Once the packages are loaded, we can begin preparing our two key inputs:

**custom_disease_data** - A simple dataset specifying the number of reported cases in each year. This can be used to seed the model or replicate past trends.

**custom_disease_parameters** - A table of disease-specific parameters that modify natural history, severity, and immunity assumptions. You can choose to supply only a subset of parameters and rely on defaults for the rest. See the table below for more detail.

| Parameter                              | Model variable                  | Description                                            | What modifying it might do                           |
|-----------------------------------------|---------------------------------|--------------------------------------------------------|------------------------------------------------------|
| incubation period                       | incubation_rate                 | Time from infection to symptoms (1/incubation_rate)    | Shorter = earlier outbreaks, longer = delays spread  |
| infectious period                       | recovery_rate                   | Duration infectious (1/recovery_rate)                  | Longer = more spread, shorter = less spread          |
| natural immunity waning period          | natural_immunity_waning         | Duration of natural immunity (1/waning rate)           | Shorter = more susceptibles over time                |
| proportion of cases that are severe     | prop_severe                     | Fraction of infections that are severe                 | Higher = more severe cases and resource use          |
| cfr for standard cases                  | cfr_normal                      | Death rate for standard (mild) cases                   | Higher = more deaths in mild cases                   |
| cfr for severe cases                    | cfr_severe                      | Death rate for severe cases                            | Higher = more deaths in severe cases                 |
| proportion of severe with complications | prop_complications              | Fraction of severe cases with complications            | Higher = more with complications after infection     |
| maternal protection (vaccine)           | protection_weight_vacc           | Proportion protected from vaccinated mothers           | Higher = more newborns protected from birth          |
| maternal protection (infection)         | protection_weight_rec            | Proportion protected from previously infected mothers  | Higher = more newborns protected from birth          |
| age maternal protection ends            | age_maternal_protection_ends     | Age when maternal protection ends                      | Higher = longer newborns are protected               |

### Custom disease data

The object `custom_disease_data` is used to provide an exact match to WHO reported cases from 1980 (the earliest data is available) in order to replicate trends in immunity seen at the current time point.

This data is also used as a switch, where if `WHO_seed_switch = TRUE`, then dynamic disease transmission is turned off at the time of the first data point, and the cases that occur in the model are provided by this dataset.

To create our custom data.frame, we need to specify two columns

**year** - The year the cases occurred in.

**cases** - The number of cases in the year.

```{r, eval = T}
custom_cases <- data.frame(
 year = c(1970, 1990, 2010),
 cases = c(1, 100, 20000000)
)
```

### Custom parameters

When customising our disease parameters, we can customise as many or as few as we want. If parameters are not provided by our custom_disease_parameters object, then it will use the defaults. For default values, see `PREVAIL::disease_parameters`.

To specify our parameters, we need to provide three arguments:

**parameter** - The parameter we are interested in.

**value** - The value we want for the parameter.

**unit** - The unit the value is provided in.

```{r, eval = T}
custom_parameters <- data.frame(
  parameter = c("incubation period", "infectious period"),
  value = c(5, 30),
  unit = c("days", "days")
)
```

All other parameters not explicitly included will be drawn from the default values.

### R0

The basic reproductive number, R0, is input as an argument in both `load_data_process_wrapper()` and `custom_data_process_wrapper()`, where it is specified as `R0 = `.


## 2. Process Custom Disease Parameters

Now that we’ve defined our disease-specific inputs—including a time series of reported cases and a custom set of parameter values—we need to convert them into a model-ready format. This is done using the `custom_data_process_wrapper()` function.

This function is the main entry point for preparing all input data required by the PREVAIL model. It combines:

- **Demographic data** for the specified country (identified by a 3-letter ISO code)
- **Vaccination coverage and schedules** using WHO datasets
- **Disease-specific parameters**, including any user-defined custom overrides
- **Case time series**, which can be used to seed or constrain the model

If any of these components are not provided, PREVAIL will automatically fall back on built-in default values. This makes it easy to customize just one part of the input while keeping the rest unchanged.

We run the wrapper like this:

```{r, eval = FALSE}
custom_params <- custom_data_process_wrapper(
  iso = "PSE",                        # ISO3 code for the country of interest
  disease = "measles",               # Disease to simulate
  R0 = 15,                           # Basic reproduction number
  custom_disease_data = custom_cases,              # Custom case time series
  custom_disease_parameters = custom_parameters    # Custom parameter values
)
```

This returns a named list of parameters that are ready to be used by the simulation engine. These include time-varying arrays for population size, contact rates, seeding events, and all disease-specific dynamics.

We can now simulate the model using `run_model_unpack_results()`:

```{r, eval = FALSE}
model_run_custom <- run_model_unpack_results(
  params = custom_params,
  simulation_length = (custom_params$input_data$year_end - custom_params$input_data$year_start) * 365,
  no_runs = 1
)
```

- `simulation_length` defines how long the model should run (in days). Here we calculate it based on the duration of the input time series.
- `no_runs` specifies the number of stochastic simulations to perform. A value of `1` is sufficient for demonstration purposes.

Finally, we can generate output plots using the `summary_plots()` function:

```{r, eval = FALSE}
plots <- summary_plots(model_run_custom, custom_params)
```

This will return a list of `ggplot2` objects that summarise the key outputs from the simulation, including time series of infections and cases, compartment sizes, and breakdowns by age where relevant.
