---
title: "disease"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{disease}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(dplyr)

options(scipen = 999)
```

# 1. Prepare your custom data

Firstly we need to load our required packages, we're using the **pacman** package here, which will also install these packages if missing.

```{r, eval = FALSE}
#If pacman is not already present, install
if(!require("pacman")) install.packages("pacman")

#Install missing packages, and load 
pacman::p_load(
  PREVAIL,
  tidyverse
)
```

Once this is done, we can start thinking about the data we would like to customise. Here we are going to be focusing on disease associated parameters. These are:

**custom_disease_data** - Case data to match.

**custom_disease_parameters** - Parameters to modify disease transmission, see the table below for more detail.

| parameter                              | model variable                  | description                                            | what modifying it might do                           |
|-----------------------------------------|---------------------------------|--------------------------------------------------------|------------------------------------------------------|
| incubation period                       | incubation_rate                 | Time from infection to symptoms (1/incubation_rate)    | Shorter = earlier outbreaks, longer = delays spread  |
| infectious period                       | recovery_rate                   | Duration infectious (1/recovery_rate)                  | Longer = more spread, shorter = less spread          |
| R0                                      | R0                             | Basic reproduction number                              | Higher = larger outbreaks, lower = smaller           |
| natural immunity waning period          | natural_immunity_waning         | Duration of natural immunity (1/waning rate)           | Shorter = more susceptibles over time                |
| proportion of cases that are severe     | prop_severe                     | Fraction of infections that are severe                 | Higher = more severe cases and resource use          |
| cfr for standard cases                  | cfr_normal                      | Death rate for standard (mild) cases                   | Higher = more deaths in mild cases                   |
| cfr for severe cases                    | cfr_severe                      | Death rate for severe cases                            | Higher = more deaths in severe cases                 |
| proportion of severe with complications | prop_complications              | Fraction of severe cases with complications            | Higher = more with complications after infection     |
| maternal protection (vaccine)           | protection_weight_vacc           | Proportion protected from vaccinated mothers           | Higher = more newborns protected from birth          |
| maternal protection (infection)         | protection_weight_rec            | Proportion protected from previously infected mothers  | Higher = more newborns protected from birth          |
| age maternal protection ends            | age_maternal_protection_ends     | Age when maternal protection ends                      | Higher = longer newborns are protected               |





