---
title: "Disease"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{disease}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(dplyr)

options(scipen = 999)
```

# 1. Prepare your custom data

Firstly we need to load our required packages, we're using the **pacman** package here, which will also install these packages if missing.

```{r, eval = FALSE}
#If pacman is not already present, install
if(!require("pacman")) install.packages("pacman")

#Install missing packages, and load 
pacman::p_load(
  PREVAIL,
  tidyverse
)
```

Once this is done, we can start thinking about the data we would like to customise. Here we are going to be focusing on disease associated parameters. These are:

**custom_disease_data** - Cases reported over time.

**custom_disease_parameters** - Parameters to modify disease transmission, see the table below for more detail.

| Parameter                              | Model variable                  | Description                                            | What modifying it might do                           |
|-----------------------------------------|---------------------------------|--------------------------------------------------------|------------------------------------------------------|
| incubation period                       | incubation_rate                 | Time from infection to symptoms (1/incubation_rate)    | Shorter = earlier outbreaks, longer = delays spread  |
| infectious period                       | recovery_rate                   | Duration infectious (1/recovery_rate)                  | Longer = more spread, shorter = less spread          |
| natural immunity waning period          | natural_immunity_waning         | Duration of natural immunity (1/waning rate)           | Shorter = more susceptibles over time                |
| proportion of cases that are severe     | prop_severe                     | Fraction of infections that are severe                 | Higher = more severe cases and resource use          |
| cfr for standard cases                  | cfr_normal                      | Death rate for standard (mild) cases                   | Higher = more deaths in mild cases                   |
| cfr for severe cases                    | cfr_severe                      | Death rate for severe cases                            | Higher = more deaths in severe cases                 |
| proportion of severe with complications | prop_complications              | Fraction of severe cases with complications            | Higher = more with complications after infection     |
| maternal protection (vaccine)           | protection_weight_vacc           | Proportion protected from vaccinated mothers           | Higher = more newborns protected from birth          |
| maternal protection (infection)         | protection_weight_rec            | Proportion protected from previously infected mothers  | Higher = more newborns protected from birth          |
| age maternal protection ends            | age_maternal_protection_ends     | Age when maternal protection ends                      | Higher = longer newborns are protected               |

### Custom disease data

The object `custom_disease_data` is used to provide an exact match to WHO reported cases from 1980 (the earliest data is available) in order to replicate trends in immunity seen at the current time point.

This data is also used as a switch, where if `WHO_seed_switch = TRUE`, then dynamic disease transmission is turned off at the time of the first data point, and the cases that occur in the model are provided by this dataset.

To create our custom data.frame, we need to specify two columns

**year** - The year the cases occured in.

**cases** - The number of cases in the year.

```{r, eval = T}
custom_cases <- data.frame(
 year = c(1990, 2010),
 cases = c(100, 2000)
)
```

### Custom parameters

When customising our disease parameters, we can customise as many or as few as we want. If parameters are not provided by our custom_disease_parameters object, then it will use the defaults. For default values, see `PREVAIL::disease_parameters`.

To specify our parameters, we need to provide three arguments:

**parameter** - The parameter we are interested in.

**value** - The value we want for the parameter.

**unit** - The unit the value is provided in.

```{r, eval = T}
custom_parameters <- data.frame(
  parameter = c("incubation period", "infectious period"),
  parameter = c(5, 30),
  unit = c("days", "days")
)
```




